---
title: "Modeling thermal trade-offs versus trade-ups in a Resistance-vs-Recovery framework"
author: "Hermina"
Date: "2025-02-05"
output:
  html_document:
    df_print: paged
    code_folding: hide
    toc: true
    toc_float:
      smooth_scroll: false
  pdf_document:
    toc: true
---

# Motivation

As summarized in the introduction of [Logan et al., 2019](https://onlinelibrary.wiley.com/doi/full/10.1111/evo.13862), there's 2 main hypotheses for thermal performance: thermodynamic effect (hereafter called "hotter is better") and specialist-generalist trade-off. The 2 are not mutually exclusive. Both hypotheses describe how the full thermal performance curve of 2 (or more) species are related.

In our experiment, we are **not** interested in the full thermal performance curve. We're only interested in what happens at the very extreme end of temperature. So we can summarize the effect of interest under a null hypothesis ("coupling") and two alternative "decoupling" hypotheses in the following figure:

```{r, trade_schematic}
plot_muSchematic <- function(mu1_ambient, mu1_extreme, mu2_ambient, mu2_extreme,
                             plot_title){
  plot(x=1:2, y=c(mu1_ambient, mu1_extreme),
       type = "n", xaxt = "n", ylim = c(-1, 2), yaxt = "n",
       xlab = "Temperature", ylab = "Growth rate", main=plot_title)
  # Add custom x-axis labels
  axis(1, at = c(1, 2), labels = c("Ambient", "Extreme"))
  # Draw the two lines
  lines(x=1:2, y=c(mu1_ambient, mu1_extreme), col="deepskyblue", lwd = 2)
  lines(x=1:2, y=c(mu2_ambient, mu2_extreme), col="coral", lwd = 2)
}


# make plot for no effect:
plot_muSchematic(mu1_ambient = 0.5, mu1_extreme = -0.5, mu2_ambient = 1.5, mu2_extreme = 0.5,
                 plot_title = "Null hypothesis: predict coupling of resistance & recovery")

# make plot for Trade-Ups:
plot_muSchematic(mu1_ambient = 1.3, mu1_extreme = -0.5, mu2_ambient = 1.5, mu2_extreme = 1,
                 plot_title = "Trade-ups (hotter is better): predict + decoupling")

# make plot for Trade-Offs:
plot_muSchematic(mu1_ambient = 0.75, mu1_extreme = 0.25, mu2_ambient = 1.75, mu2_extreme = -0.75,
                 plot_title = "Trade-offs: predict - decoupling")
```

The question is: how do these different hypotheses map onto Maddy's Resistance-Recovery framework? And what do we expect to see under each of these hypotheses as we increase the duration of the heat pulse event? My intuition is that:

1. when the 2 lines are parallel (i.e., the null hypothesis plot above) then recovery $=$ resistance so there will be a *coupling* of these responses;

2. when there are trade-ups (i.e., hotter is better hypothesis) then recovery $>$ resistance so there will be a "+ decoupling" of resistance from recovery that is *above* the resistance = recovery line;

3. but that when there are trade-offs then recovery $<$ resistance: so there will be a "- decoupling" of resistance from recovery that is *below* the resistance = recovery line.

# The Model

I think it is sufficient to consider a model of Logistic growth with multiple species $i$, where each species has population size $n_i$: $\frac{1}{n_i} \frac{dn_i}{dt} = \mu_i(T)(1-\frac{\Sigma n_i}{K})$.

Growth rate is an explicit function of temperature ($\mu_i(T)$) but as we have a pulse perturbation, we're actually only interested in two growth rates for each species: that at ambient temperature and at the extreme temperature. The ambient growth rates we have already measured so that's fine. The trouble is that the growth rates are negative for extreme temperatures but we cannot measure death... Anyway we can still eye-ball things.

Finally, we assume that carrying capacity is temperature and species independent. We can estimate the carrying capacity from the control treatment because the temperature stays constant at its ambient value.

For starters, we have just 2 species in the community.


19 Dec Notes: I think it's worthwhile to simplify the model even further (for now) by considering only exponential growth. That's because we have no antagonistic interactions in the model, it's just competition on density-dependent growth rate. So we can get an idea of how the batch culture results will look like by simulating less time for exponential growth. Then I can more quickly test if my intuition about resistance/recovery make sense before I bother to specify the batch culture model.

## Building the model

```{r, define_model}
library(deSolve) # for solving ordinary differential equations
library(tidyverse) # for everything

# define the set of ordinary differential equations
ODE_2sp <- function(time, init, params){
  with (as.list(c(time, init, params)), {
    dN1_dt = N1 * mu1 * (1 - (N1 + N2)/K)
    dN2_dt = N2 * mu2 * (1 - (N1 + N2)/K)
    return(list(c(dN1 = dN1_dt, dN2 = dN2_dt)))
  })
}

# a function to simulate the community for a constant temperature
sim_1temp <- function(time, init, params) {
  # solve the system of ODE's for the requested time points:
  sim <- ode(func = ODE_2sp, y = init, parms = params, times = time, method="ode23")
  # make a long data.frame with trees and grass over time
  return(data.frame(sim))
}

# a function to simulate a pulse perturbation
sim_pulse <- function(times, # a vector of length 3: when extreme heat starts, when it stops, and total time
                      mu_ambient, # a named vector of length 2: mu1 and m2
                      mu_extreme, # a named vector of length 2: mu1 and m2
                      K,
                      init # a named vector of length 2: N1 and N2
                      ) {
  # simulate the time at ambient temperature before the start of the heat pulse
  sim <- sim_1temp(time = 0:times[1],
                   init = init,
                   params = c(K=K, mu_ambient["mu1"], mu_ambient["mu2"]))
  
  # then simulate the heat pulse itself and append it to the simulation results
  sim <- rbind(sim[-(times[1]+1),], # remove final value of previous simulation bc it's redundant
               sim_1temp(time = times[1]:times[2], # from beginning to end of heat pulse
                         init = unlist(sim[(times[1]+1),-1]), # initialize from the final values of the previous simulation
                         params = c(K=K, mu_extreme["mu1"], mu_extreme["mu2"])) # use the growth rates during extreme heat
               )
  
  # finally simulate the recovery after heat pulse and append it to the results
  sim <- rbind(sim[-(times[2]+1),],
               sim_1temp(time = times[2]:times[3], # from end of heat pulse to end of time
                         init = unlist(sim[(times[2]+1),-1]),
                         params = c(K=K, mu_ambient["mu1"], mu_ambient["mu2"])) # growth rates at ambient
               )
  
  return(sim)
}

# TO DO: a function to simulate serial transfers with/out pulse perturbations
sim_serialTransf <- function(trmt) {
  # check that trmt has an allowed value
  stopifnot(trmt %in% c("ctrl", "6h", "12h", "24h", "48h"))
  
  # 
}




##################################################

plot_muSchematic(mu1_ambient = 0.76, mu1_extreme = 0.7, mu2_ambient = 0.97, mu2_extreme = 0.6,
                 plot_title = "example")

# let's try out an example
State <- c(N1=0.001, N2=0.001)

# let's plot the example
simex <- sim_pulse(times=c(5,18,24), init=State, K=10^6,
                   mu_ambient=c(mu1=0.76, mu2=0.97), mu_extreme=c(mu1=0.7, mu2=0.6)) %>%
          pivot_longer(cols = N1:N2)

ggplot(simex, aes(x=time, y=value, colour=name)) +
  geom_line() +
  scale_y_log10()
```


